---
title: "SEA-AD_PFC"
format: html
editor: visual
---

### Set up

```{r}
source("configs/DLPFC_Config.R")
source('utility_scripts/seurat_functions.R')
```

## Analysis Workflow

The goal of this workflow is to perform a CD8+ T cell abundance analysis to assess whether there are an increased proportion of Cd8+ T cells in patients with dementia.

### Reading in the data

```{r}
immune_cells<-readRDS(paste0(PATH_TO_SEURAT_OBJ,".rds"))
```

### Fix Meta data

```{r}
colnames(immune_cells@meta.data) <- make.names(colnames(immune_cells@meta.data), unique=TRUE)
```

### Standard Seurat Clustering workflow

Following best practices outlined by this workflow: <https://satijalab.org/seurat/articles/pbmc3k_tutorial.html>

Sourced from seurat_functions.R

```{r}
set.seed(1)
immune_list<-cluster_seurat(immune_cells)
immune_list$DimPlot
```

### 

Feature plot of all immune cells using CD247 (CD3 marker that was used to identify the Lymphocyte cluster in the previous paper)

```{r fig.height= 10, fig.width=10}
cust_feature(immune_list$seurat,c("ENSG00000198821"))
```

```{r fig.height=10, fig.width=20}
VlnPlot(immune_list$seurat,assay="RNA",slot='data', features = c("ENSG00000198821"),raster=FALSE,pt.size = 0,ncol = 2) #Cluster 22
```

This pointed to cluster 22 as the lymphocyte cluster

### Lymphocyte Extraction

```{r}
#filter for clusters of interest
lymphocytes<-subset(x=immune_list$seurat,subset =seurat_clusters %in% c(22))

# Count total number of cells
lymphocytes@meta.data %>% nrow()# 644

# Count the total number of donors
length(unique(lymphocytes@meta.data$donor_id)) # 77
```

### Subcluster Lymphocytes

```{r fig.width=6,fig.height= 4}
seurat<-lymphocytes
set.seed(1)
seurat <- NormalizeData(object = seurat)
seurat <- FindVariableFeatures(object = seurat,selection.method = "vst", nfeatures = 2000)
seurat <- ScaleData(object = seurat)
seurat <- RunPCA(object = seurat)
seurat <- FindNeighbors(object = seurat, dims = 1:20)
seurat <- FindClusters(object = seurat, resolution = 1.2)
seurat <- RunUMAP(object = seurat, dims = 1:20)
plot<-DimPlot(object = seurat, reduction = "umap",label=T,label.size = 5)
lympho_list<-list(seurat=seurat,
               DimPlot=plot)
lympho_list$DimPlot

ggsave("../output/figure_3/panel_3_d.tiff", plot = lympho_list$DimPlot,
       width = 6, height = 4, units = "in", dpi = 600,
       device = "tiff", compression = "lzw")
```

### Filtering Lymphocytes

```{r fig.height=4, fig.width=6}
library(patchwork)  # for combining plots

# Ensembl IDs and gene symbols
features <- c("ENSG00000010610","ENSG00000153563","ENSG00000172116",
              "ENSG00000167286","ENSG00000198851", 
              "ENSG00000149294")

gene_symbols <- c("CD4", "CD8A", "CD8B", "CD3D", "CD3E", "NCAM1")

# Create the list of plots
plots <- VlnPlot(lympho_list$seurat, features = features, pt.size = 0.1)

# Rename titles using gene symbols
for (i in seq_along(plots)) {
  plots[[i]] <- plots[[i]] + ggtitle(gene_symbols[i])
}

# Combine into one figure
combined_plot <- wrap_plots(plots)

# Display

ggsave("../output/figure_3/panel_3_e.tiff", plot = combined_plot,
       width = 6, height = 4, units = "in", dpi = 600,
       device = "tiff", compression = "lzw")
combined_plot
```


Checking number of cells and donors in each cluster

```{r}
lympho_list$seurat@meta.data %>% group_by(seurat_clusters) %>%summarise(n=n(),unique_donors=length(base::unique(donor_id))) # 0 is NK, #
```

### Assign Cluster Identities

```{r}
# Recode Cluster names
lympho_list$seurat@meta.data$seurat_clusters <- recode(lympho_list$seurat@meta.data$seurat_clusters %>% as.factor(), `0` = "NK", `1` = "CD8+",
                                                      `2`= "CD8+",
                                                     `3`="CD8+",
                                                    `4`="CD8+",
                                                     `5`="CD8+",
                                                     `6`="CD4+",
                                                    `7`="Mystery")
```

### Set up for abundance test

```{r}
# Getting Subclusters at the individual level
individual_sub_clust<-lympho_list$seurat@meta.data %>% group_by(donor_id,seurat_clusters) %>% summarise(count=n())
individual_total_cells<-immune_list$seurat@meta.data %>% group_by(donor_id) %>% summarise(count=n())

# Pivot to column format
individual_sub_clust<-pivot_wider(individual_sub_clust, id_cols = donor_id,values_from = count,names_from = seurat_clusters)

# Merge with clinical meta data on idividaul basis
# Pull invidual meta data
clinical<-lympho_list$seurat@meta.data %>% dplyr::select(c("donor_id","PMI","sex","Cognitive.status","Age.at.death")) %>% distinct()
#Merge
clinical<- merge(clinical,individual_sub_clust,by="donor_id")
# Get total cells
clinical$total_cells<-individual_total_cells$count[match(clinical$donor_id,individual_total_cells$donor_id)]

# Add new columns
clinical<-clinical %>% mutate(norm_CD8= `CD8+`/total_cells,norm_NK= `NK`/total_cells,norm_CD4= `CD4+`/total_cells)
```

### Abundance Analysis

```{r}
cd8<-Abudance_analysis(clinical, column = "norm_CD8",log_prop =  FALSE,covariates =  COVARIATES, disease = "Cognitive.status")
```

```{r}
cd8$wilcox_test
```

```{r}
cd8$chi_squared_test
```

```{r}
cd8$model %>% summary()
```

```{r}
plot2 <- ggplot() +
        geom_violin(data = clinical, aes(y = norm_CD8,x= Cognitive.status, fill = Cognitive.status)) + geom_jitter (data = clinical, aes(y = norm_CD8,x= Cognitive.status, fill = Cognitive.status),width = 0.1, size=0.5)+
         labs(y = "CD8+ T cells (fraction of total immune cells)",
         x = NULL) + theme_classic()+ scale_fill_manual(values=c("#999999", "#6829e7"))+ theme(legend.position="none",axis.text.x =element_text(size=14,face="bold"),axis.title=element_text(size=15))
ggsave("../output/figure_3/panel_3_f_clinical                                                  .tiff", plot = plot2,
       width = 6, height = 6, units = "in", dpi = 600,
       device = "tiff", compression = "lzw")         
plot2
```

### APOE4 Analysis

```{r}
multivar<-glm(as.formula(paste0("norm_CD8","~PMI+sex+Cognitive.status+Age.at.death+APOE4.status")),data=clinical)
summary(multivar)
```

### Table one
```{r}
library(tableone)
clinical$Cognitive.status<-droplevels(clinical$Cognitive.status)
filt_tab=clinical%>% filter(`CD8+`>0)
## Vector of variables to summarize
myVars <- c("sex","PMI","Age.at.death","APOE4.status")
## Vector of categorical variables that need transformation
catVars <- c("sex","PMI","Age.at.death","APOE4.status")
## Create a TableOne object
tab2 <- CreateTableOne(vars = myVars, data = filt_tab, factorVars = catVars,strata="Cognitive.status")
tab2
```

```{r}
tab2Mat <- print(tab2, exact = "stage", quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
## Save to a CSV file
write.csv(tab2Mat, file = "../output/supplementary_tables/SEA_PFC_Table.csv")
```
